Headgroup structure and cation binding in phosphatidylserine lipid bilayers

### 2019-05-27 Manuscript/manuscriptPS.tex
Update text based on comments by Markus Miettinen delivered by email

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPS.pdf}
  \caption{\label{HGorderParametersPS}
    Order parameters of PS headgroup ($\beta$ and $\alpha$) and
    glycerol backbone (g$_3$, g$_2$, g$_1$) from NMR experiments (horizontal lines),
    and MD simulations with different force fields (symbols).
    Experimental data for DOPS are measured with 0.1~M of NaCl~\cite{browning80},
    while all the other data are without additional salt.
    The data for DOPS is at 303~K and the data for POPS is at 298~K.
    Light blue areas span 0.04 units around the average of the extremal experimental values,
    in accordance with the expected quantitative accuracy of experiments~\cite{ollila16}.
    The vertical bars shown for all simulation values (excl. MacRog~K$^+$)
    are not error bars, but demonstrate that for these systems
    we had at least two data sets; the ends of the bars mark the extreme values
    from the sets, and the symbol marks their measurement-time-weighted average.
  }
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/structures.eps}
  \caption{\label{HGstructuresPSandPC}
    Overlayed snapshots from simulations conducted with CHARMM36---the force field producing the best agreement with experiments---demonstrate the conformational fluctuations around
    (A) C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ and  O$_\alpha$-C$_\alpha$-C$_\beta$-N
    of PS headgroup and (B) N-C$_\beta$-C$_\alpha$-O$_\alpha$ and C$_\beta$-C$_\alpha$-O$_\alpha$-P
    dihedrals of PS and PC headgroups. The trajectory used for CHARMM36 POPC is available at~\citenum{POPCcharmm36T303K}.
  }

@@

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPS}
    Changes of POPC headgroup order parameters with increasing amount of POPS in POPC:POPS mixtures at 298~K.
    Experimental values are from Ref. \citenum{scherer87} with the signs measured in Ref.~\citenum{ferreira16}.
    The values from the CHARMM36 and GROMOS-CKP simulations with sodium are averages of two independent simulations and
    the error bar is given as the difference between the results divided by two.
    % {\color{red} Why divided by two?}
    % Because then both values fit within the error bar.
  }
\end{figure}

@@

Membranes containing PS lipids are always accompanied with counterions that
modulate electrostatic interactions between lipids and other biomolecules. MD simulations have suggested
that counterions reduce the area per lipid of PS bilayers compared to PC
bilayers~\cite{pandit02,mukhopadhyay04,pedersen06} by screening the repulsion between charged lipid headgroups. We explore this by quantifying the counterion density profiles along the membrane normal accompanied by the areas per lipid in Fig.~\ref{NAdensPOPS}.
The force fields studied show significant differences in both binding affinity
and distribution of ions at the interface.
The experimental area per lipid (62.7~$\pm$~1.3~\AA$^2$) \cite{pan14} 
is reproduced only in GROMOS-CKP and in the MacRog simulation
with potassium counterions, while other models give considerably smaller values (Fig.~\ref{NAdensPOPS}).
However, the counterion binding and the concomitant electrostatic screening of the headgroup
repulsion does not fully explain the low area per molecule values
since the MacRog simulation, which has the strongest sodium binding
(the lowest concentrations in bulk water), gives the same area per molecule
as the CHARMM36-UA simulation, which has significantly weaker counterion binding
affinity. On the other hand, MacRog simulations with potassium produce a larger area per molecule (63~\AA$^2$) than
with sodium (53~\AA$^2$) in line with the weaker potassium binding affinity (Fig.~\ref{NAdensPOPS}).
The results are consistent with the previous study suggesting that the
low areas per molecule in PS lipid bilayers originate from the combination
of both the counterion binding and intermolecular interactions between lipid headgroups~\cite{petrache04}.
% Would be potentially interesting to flesh out what the role of the hydrating water molecules play in the
% specific lipid headgroup/ion configurations that are sampled in simulations. My feeling is that "fully hydrated"
% lipids might carry with them differing amounts of water between the different FFs.. -JJM

@@

The experimentally observed modulation of headgroup order parameters
by increasing salt concentration (the molecular electrometer concept) has been previously used to evaluate the cation binding to zwitterionic PC bilayers in simulations \cite{catte16}.
Studying binding of cations to negatively charged lipid bilayers is less straighforward due to the presence of cationic counterions (the lack of a ion-free reference state) and the analysis is further complicated by the artificial aggregation of counterions
observed in some simulations (section \ref{mixtureTOadditionalCIs} in the SI)
% \todo{Why isn't this a problem with PC}.
% SAMULI: The difference between ion-free state (present in PC) and low amount of monovalent ions is larger than
% between counterion only and low amount of monovalent ions in charged lipids. I believe that therefore we can
% make conclusions from smaller ion concentrations without aggregation in PC lipids, but not in charged lipids.
Therefore, we evaluate here the amount of bound charge not by adding salt
(although this is discussed in the SI section \ref{mixtureTOadditionalCIs}),
but by studying the changes of the headgroup order parameters with increasing amount of
negatively charged lipids (and thus increasing amount of cationic counterions) in the bilayer.

Experimentally, the both headgroup order parameters of POPC
increase when negatively charged POPS lipids are incorporated in lipid bilayer
(section \ref{HGorderparametersPCvsPEPSPGchol})~\cite{seelig87,scherer87}.
This is reproduced in the MacRog simulations with potassium counterions (Fig. \ref{HGorderparametersPCvsPS}),
which have the weakest binding affinity to POPS lipid bilayers (Fig.~\ref{NAdensPOPS}).
The CHARMM36, Berger and GROMOS-CKP simulations either exhibit no change or show a decrease
in either or both the POPC headgroup order parameters as the amount of POPS increases (Fig. \ref{HGorderparametersPCvsPS}).
Therein, the stronger counterion binding cancels
the effect of negatively charged headgroups and prevents the experimentally observed
increase of headgroup order parameters with growing amount of PS lipids.
Therefore, we suggest that the relatively weak binding of potassium
in the MacRog simulations (Fig. \ref{NAdensPOPS}) produces the most
realistic surface charge density in membranes containing PS lipids,
while the other tested models overestimate the counterion
binding affinity. The results are consistent with the behavior of headgroup order
parameters as a function of added counterions analyzed in section \ref{mixtureTOadditionalCIs}
in the SI.

The reduced forking of the POPS $\alpha$-carbon (Fig.~\ref{HGorderparametersPSvsPC})
together with other experimental results suggest that the PS headgroup structure becomes less rigid when diluted with
POPC~\cite{browning80,buldt81,roux90,roux91,scherer87}.
Unfortunately, none of the tested models correctly reproduce the modulation of POPS headgroup order
parameters with increasing amount of POPC in POPC:POPS mixtures (Fig. \ref{HGorderparametersPSvsPC})
and more accurate force fields are needed
to correctly describe the PC--PS headgroup interactions in MD simulations.
%Having this conclusion in every paragraph gets a bit repetive

@@

Experimentally, the POPS headgroup order parameters in POPC:POPS (5:1) mixtures
exhibit a strong dependence of CaCl$_2$ at small concentrations yet they saturate already at a concentration below 100~mM (Fig. \ref{changesWITHCaClPS}).
The $\beta$-carbon order parameter increases with added CaCl$_2$,
whereas the larger $\alpha$-carbon order parameter decreases. Moreover, a slight increase is observed in
the smaller $\alpha$-carbon value. All these changes are significantly overestimated in the
tested simulation models---including CHARMM36---which underestimated the binding affinity.
Importantly, the different simulation models predict qualitatively different behavior
for the two POPS $\alpha$-carbon order parameters as a function of added calcium.
For example, both order parameters decrease in Berger, but increase
in MacRog, whereas Lipid14/17 and CHARMM36 models exhibit more complex behavior.
This is in contrast to the PC headgroup, where a
qualitatively correct response to bound ions is reproduced
by all simulation models, despite the significant discrepancies in the headgroup
structures observed in salt-free simulations~\cite{catte16}.
The divergent response of the Berger model may arise from the ring like structures
observed in the headgroup region in this model (Fig. 6 in Ref. \citenum{mukhopadhyay04}).
%These ring like structures are a widespread feature of typical Berger based lipid force
%fields containing explicit hydrogen atoms in the head group
%[REFS https://pubs.acs.org/doi/abs/10.1021/jp900645z and https://www.sciencedirect.com/science/article/pii/S0300908408000692 and https://pubs.acs.org/doi/abs/10.1021/jp9110019]
Therefore, we conclude that
the improvement of force fields is necessary to correctly capture the interactions between the
PS headgroup and calcium ions in MD simulations.

@@

%Lipids with PS headgroups, and their interactions with ions, play an important role in lipid-mediated signaling processes \cite{leventis10,yeung08}. Attemps to use MD simulations to interpret the spectroscopic data have produced contradictory results for the calcium binding details to PSheadgroups \cite{melcrova16,valentine18,hallock18}.  
%Here (as was previously done for PC lipids \cite{botan15,catte16})
Here, we used the headgroup C--H bond order parameters and the open collaboration approach to evaluate the quality
of the headgroup structure and the ion binding affinity 
in available MD models of PS.
The main advantage of this approach is the direct connection
between the order parameters measured accurately in experiments and calculated readily from simulations. 
This reduces the ambiguity in the interpretation of experiments.

First, we complemented the available experimental information \cite{browning80,roux90} by measuring the signs of the PS headgroup order parameters,
and then proceeded to compare MD simulation results from several force fields with the experimental data.
This revealed that none of the force fields
%tested using the NMRlipids open collaboration
reproduce the PS headgroup order parameters within the experimental accuracy. However,
the best models for the serine headgroup suggested a characteristic rigid conformation for its
carboxyl group. Comparison to the experimentally observed order parameters in POPC:POPS (5:1) bilayers  at varying ion
concentrations \cite{roux90} then showed that the tested MD force fields
overestimate the cation binding affinity to these bilayers with two exceptions: 1) the MacRog simulation with potassium counterions appears to produce the most realistic monovalent ion binding
affinity to PS-containing lipid bilayers, and 2) the CHARMM36 force field with the recently introduced
NBfix correction for calcium \cite{kim16} underestimated the calcium binding affinity.
The experimentally measured response of the PS headgroup order parameter to the bound calcium, and to the dilution of bilayer with zwitterionic PC lipids, were not
qualitatively reproduced in any of the tested force fields, which underlines the need for more accurate
MD force fields to study the biological function of PS lipids.
This is in contrast to the previous results with PC lipids,
where the experimentally measured headgroup order parameter responses to the bound charge
were in qualitative agreement even though the headgroup structures themselves were
incorrect and the cation binding affinities were overestimated \cite{catte16}.

We expect our results to pave the way for the development of better MD force fields
for PS lipids. The quality of conformational ensembles produced by the various models can be evaluated based on the headgroup order parameters, which we hope to guide the development of force fields towards models that correctly reproduce the PS headgroup structures. The cation binding in the force fields can be improved based on the experimental headgroup order parameter data from POPC:POPS (5:1) mixtures under different cation concentrations. This was already demonstrated for POPC using the electronic continuum correction \cite{melcr18}, and a similar study for POPS is being progressed separately \cite{ECCpops}.

>>>>

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPS.pdf}
  \caption{\label{HGorderParametersPS}
    C--H bond order parameters, $S_{\rm CH}$, of PS headgroup ($\beta$ and $\alpha$) and
    glycerol backbone (g$_3$, g$_2$, g$_1$) carbons from NMR experiments (horizontal lines),
    and MD simulations with different force fields (symbols).
    Experimental data for DOPS are measured with 0.1~M of NaCl~\cite{browning80},
    while all the other data are with counterions only.
    The data for DOPS is at 303~K and for POPS at 298~K.
    Light blue areas span 0.04 units around the average of the extremal experimental values,
    in accordance with the expected quantitative accuracy of experiments~\cite{ollila16}.
    The vertical bars shown for all simulation values (excl. MacRog~K$^+$)
    are not error bars, but demonstrate that for these systems
    we had at least two data sets; the ends of the bars mark the extreme values
    from the sets, and the symbol marks their measurement-time-weighted average.
  }
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/structures.eps}
  \caption{\label{HGstructuresPSandPC}
    Overlayed snapshots from simulations conducted with CHARMM36---the force field producing the best agreement with experiments---demonstrate the conformational fluctuations around
    (A) C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ and  O$_\alpha$-C$_\alpha$-C$_\beta$-N dihedrals
    of the PS headgroup and (B) N-C$_\beta$-C$_\alpha$-O$_\alpha$ and C$_\beta$-C$_\alpha$-O$_\alpha$-P
    dihedrals of the PS and PC headgroups. The trajectory used for CHARMM36 POPC is available at Ref.~\citenum{POPCcharmm36T303K}.
  }

@@

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPS}
    Changes of POPC headgroup order parameters with increasing amount of POPS in POPC:POPS mixtures at 298~K.
    Experimental values are from Ref. \citenum{scherer87}, with the signs measured in Ref.~\citenum{ferreira16}.
    The values from the CHARMM36 and GROMOS-CKP simulations with sodium are averages of two independent simulations and
    the error bar is given as the difference between the results divided by two.
    % {\color{red} Why divided by two?}
    % Because then both values fit within the error bar.
  }
\end{figure}

@@

Membranes containing PS lipids are always accompanied with counterions that
modulate electrostatic interactions between the lipids and other biomolecules. MD simulations have suggested
that counterions reduce the area per lipid of PS bilayers compared to PC
bilayers~\cite{pandit02,mukhopadhyay04,pedersen06} by screening the repulsion between charged lipid headgroups.
We explored this by quantifying the counterion density profiles along the membrane normal, accompanied by the areas per lipid, see Fig.~\ref{NAdensPOPS}.
The force fields studied show significant differences in both binding affinity
and distribution of ions at the interface.
The experimental area per lipid (62.7~$\pm$~1.3~\AA$^2$) \cite{pan14} 
is reproduced only in GROMOS-CKP and in the MacRog simulation
with potassium counterions, while other models give considerably smaller values (Fig.~\ref{NAdensPOPS}).
However, the counterion binding and the concomitant electrostatic screening of the headgroup
repulsion does not fully explain the low area per molecule values
since the MacRog simulation, which has the strongest sodium binding
(the lowest concentrations in bulk water), gives the same area per molecule
as the CHARMM36-UA simulation, which has significantly weaker counterion binding
affinity. On the other hand, MacRog simulations with potassium produce a larger area per molecule (63~\AA$^2$) than
with sodium (53~\AA$^2$) in line with the weaker potassium binding affinity (Fig.~\ref{NAdensPOPS}).
The results are consistent with a previous study suggesting that the
low areas per molecule in PS lipid bilayers originate from the combination
of both the counterion binding and intermolecular interactions between lipid headgroups~\cite{petrache04}.
% Would be potentially interesting to flesh out what the role of the hydrating water molecules play in the
% specific lipid headgroup/ion configurations that are sampled in simulations. My feeling is that "fully hydrated"
% lipids might carry with them differing amounts of water between the different FFs.. -JJM

@@

The experimentally observed modulation of headgroup order parameters
by increasing salt concentration (the molecular electrometer concept) has been previously used to evaluate the cation binding to zwitterionic PC bilayers in simulations \cite{catte16}.
Studying binding of cations to negatively charged lipid bilayers is less straighforward due to the presence of cationic
counterions (the lack of an ion-free reference state). The analysis is further complicated by the artificial aggregation of counterions
observed in some simulations (section \ref{mixtureTOadditionalCIs} in the SI).
% \todo{Why isn't this a problem with PC}.
% SAMULI: The difference between ion-free state (present in PC) and low amount of monovalent ions is larger than
% between counterion only and low amount of monovalent ions in charged lipids. I believe that therefore we can
% make conclusions from smaller ion concentrations without aggregation in PC lipids, but not in charged lipids.
Therefore, we evaluate here the amount of bound charge not by adding salt
(although this is discussed in the SI section \ref{mixtureTOadditionalCIs}),
but by studying the changes of the headgroup $S_{\rm CH}$ order parameters with increasing amount of
negatively charged lipids (and thus increasing amount of cationic counterions) in the bilayer.

Experimentally, the both headgroup order parameters of POPC
increase when negatively charged POPS lipids are incorporated in the bilayer
(section \ref{HGorderparametersPCvsPEPSPGchol})~\cite{seelig87,scherer87}.
This is reproduced in the MacRog simulations with potassium counterions (Fig. \ref{HGorderparametersPCvsPS}),
which have the weakest binding affinity to POPS lipid bilayers (Fig.~\ref{NAdensPOPS}).
The CHARMM36, Berger and GROMOS-CKP simulations either exhibit no change or show a decrease
in either or both the POPC headgroup order parameters as the amount of POPS increases (Fig. \ref{HGorderparametersPCvsPS}).
Therein, the stronger counterion binding cancels
the effect of negatively charged headgroups and prevents the experimentally observed
increase of headgroup order parameters with growing amount of PS lipids.
Therefore, we suggest that the relatively weak binding of potassium
in the MacRog simulations (Fig. \ref{NAdensPOPS}) produces the most
realistic surface charge density in membranes containing PS lipids,
while the other tested models overestimate the counterion
binding affinity. The results are consistent with the behavior of headgroup order
parameters as a function of added counterions analyzed in section \ref{mixtureTOadditionalCIs}
in the SI.

The reduced forking of the POPS $\alpha$-carbon (Fig.~\ref{HGorderparametersPSvsPC})
together with other experimental results suggest that the PS headgroup structure becomes less rigid when diluted with
POPC~\cite{browning80,buldt81,roux90,roux91,scherer87}.
Unfortunately, none of the tested models correctly reproduce the modulation of POPS headgroup order
parameters with increasing amount of POPC in POPC:POPS mixtures (Fig. \ref{HGorderparametersPSvsPC}).
More accurate force fields are needed
to correctly describe the PC--PS headgroup interactions in MD simulations.
%Having this conclusion in every paragraph gets a bit repetive

@@

Experimentally, the POPS headgroup order parameters in POPC:POPS (5:1) mixtures
exhibit a strong response to small concentrations of CaCl$_2$, which saturates below 100~mM (Fig. \ref{changesWITHCaClPS}).
The $\beta$-carbon order parameter increases with added CaCl$_2$,
whereas the larger $\alpha$-carbon order parameter decreases. Moreover, a slight increase is observed in
the smaller $\alpha$-carbon value. All these changes are significantly overestimated in the
tested simulation models---including CHARMM36---which underestimated the binding affinity.
Importantly, the different simulation models predict qualitatively different behavior
for the two POPS $\alpha$-carbon order parameters as a function of added calcium.
For example, both order parameters decrease in Berger, increase
in MacRog, whereas Lipid14/17 and CHARMM36 models exhibit more complex behavior.
This is in contrast to the PC headgroup, where a
qualitatively correct response to bound ions is reproduced
by all simulation models, despite the significant discrepancies in the headgroup
structures observed in salt-free simulations~\cite{catte16}.
The divergent response of the Berger model may arise from the ring like structures
observed in the headgroup region in this model (Fig. 6 in Ref. \citenum{mukhopadhyay04}).
%These ring like structures are a widespread feature of typical Berger based lipid force
%fields containing explicit hydrogen atoms in the head group
%[REFS https://pubs.acs.org/doi/abs/10.1021/jp900645z and https://www.sciencedirect.com/science/article/pii/S0300908408000692 and https://pubs.acs.org/doi/abs/10.1021/jp9110019]
Therefore, we conclude that
the improvement of force fields is necessary to correctly capture the interactions between the
PS headgroup and calcium ions in MD simulations.

@@

%Lipids with PS headgroups, and their interactions with ions, play an important role in lipid-mediated signaling processes \cite{leventis10,yeung08}. Attemps to use MD simulations to interpret the spectroscopic data have produced contradictory results for the calcium binding details to PSheadgroups \cite{melcrova16,valentine18,hallock18}.  
%Here (as was previously done for PC lipids \cite{botan15,catte16})
We used the headgroup C--H bond order parameters, $S_{\rm CH}$, and the open collaboration approach to evaluate the quality
of the headgroup structure and the ion binding affinity 
in available MD models of PS lipid.
The main advantage of this approach is the direct connection
between the order parameters measured accurately in experiments and calculated readily from simulations. 
This reduces the ambiguity in the interpretation of experiments.

First, we complemented the available experimental information \cite{browning80,roux90} by measuring the signs of the PS headgroup order parameters,
and then proceeded to compare MD simulation results from several force fields with the experimental data.
This revealed that none of the force fields
%tested using the NMRlipids open collaboration
reproduce the PS headgroup order parameters within the experimental accuracy. However,
the best models for the PS headgroup suggested a characteristic rigid conformation for its
carboxyl group. Comparison to the experimentally observed order parameters in POPC:POPS (5:1) bilayers  at varying ion
concentrations \cite{roux90} then showed that the tested MD force fields
overestimate the cation binding affinity to these bilayers with two exceptions: 1) the MacRog simulation with potassium counterions appears to produce the most realistic monovalent ion binding
affinity to PS-containing lipid bilayers, and 2) the CHARMM36 force field with the recently introduced
NBfix correction for calcium \cite{kim16} underestimates the calcium binding affinity.
The experimentally measured response of the PS headgroup order parameters to the bound calcium, as well as to the dilution of bilayer with zwitterionic PC lipids, were not
qualitatively reproduced in any of the tested force fields. This underlines the need for more accurate
MD force fields to study the biological function of PS lipids.
This is in contrast to the previous results with PC lipids,
where the experimentally measured headgroup order parameter responses to the bound charge
were in qualitative agreement even though the headgroup structures themselves were
incorrect and the cation binding affinities were overestimated \cite{catte16}.

We expect our results to pave the way for the development of better MD force fields
for PS lipids. The quality of conformational ensembles produced by the various models can be evaluated based on the headgroup order parameters, which we hope to guide the development of force fields towards models that correctly reproduce the PS headgroup structures. The cation binding can be improved based on the experimental headgroup order parameter data from POPC:POPS (5:1) mixtures under different cation concentrations. This was already demonstrated for POPC using the electronic continuum correction \cite{melcr18}, and a similar study for POPS is being progressed separately \cite{ECCpops}.
